---
title: "hw4"
author: "Yi Wang"

format:
  html:
    embed-resources: true

execute:
  warning: false
  cache: false
  freeze: auto

fig-width: 7
fig-height: 4
---

```{r}
library(microbenchmark)
library(parallel)
library(RSQLite)
library(DBI)
```

```{r}
# Total for each row
fun1 <- function(mat) {
  n <- nrow(mat)
  ans <- double(n) 
  for (i in 1:n) {
    ans[i] <- sum(mat[i, ])
  }
  ans
}

fun1alt <- function(mat) {
  rowSums(mat)
}

# running cumulative sum by row
fun2 <- function(mat) {
  n <- nrow(mat)
  k <- ncol(mat)
  ans <- mat
  for (i in 1:n) {
    for (j in 2:k) {
      ans[i,j] <- mat[i, j] + ans[i, j - 1]
    }
  }
  ans
}

fun2alt <- function(mat) {
  t(apply(mat, 1, cumsum))
}
```

# Question 1 

```{r}
# Use the data with this code
set.seed(2315)
dat <- matrix(rnorm(200 * 100), nrow = 200)
# Check equality
all.equal(fun1(dat), fun1alt(dat))
all.equal(fun2(dat), fun2alt(dat))
```

```{r}
# Test for the first
microbenchmark::microbenchmark(
  fun1(dat),
  fun1alt(dat), unit = "relative"
)
# Test for the second
microbenchmark::microbenchmark(
  fun2(dat),
  fun2alt(dat), unit = "relative"
)
```

For fun1, the original loop-based implementation is significantly slower. The benchmark shows that fun1alt(), which uses the vectorized rowSums() function, is approximately 15 times faster than the original version. This improvement results from removing the explicit R-level loop and relying on optimized internal C code. For fun2, the optimized version fun2alt() is also substantially faster. The benchmark indicates that fun2alt() achieves about a 3-fold speedup compared to the original nested-loop implementation. By using apply() with cumsum(), we leverage vectorized operations that drastically reduce overhead. Overall, the vectorized implementations provide major speed improvements while producing identical numerical results.

## Make things run faster with parallel computing

```{r}
sim_pi <- function(n = 1000, i = NULL) {
  p <- matrix(runif(n*2), ncol = 2)
  mean(rowSums(p^2) < 1) * 4
}

# Here is an example of the run
set.seed(156)
sim_pi(1000) # 3.132
```

```{r}
# This runs the simulation a 4,000 times, each with 10,000 points
set.seed(1231)
system.time({
  ans <- unlist(lapply(1:4000, sim_pi, n = 10000))
  print(mean(ans))
})
```

# Question 2

```{r}
cl <- makeCluster(4)

clusterExport(cl, varlist = c("sim_pi"))

set.seed(1231)
system.time({
  ans <- unlist(parLapply(cl, 1:4000, sim_pi, n = 10000))
  print(mean(ans))
})
stopCluster(cl)
```

The parallel version using parLapply() completed in 0.255 seconds, whereas the serial lapply() version required 0.706 seconds.

## SQL

```{r}
# install.packages(c("RSQLite", "DBI"))

library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Download tables
film <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/film.csv")
film_category <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/film_category.csv")
category <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/category.csv")

# Copy data.frames to database
dbWriteTable(con, "film", film)
dbWriteTable(con, "film_category", film_category)
dbWriteTable(con, "category", category)

dbListTables(con) 
```

# **Question 3**

```{sql, connection=con}
SELECT
  rating,
  COUNT(*) AS n_films
FROM film
GROUP BY rating;
```

# **Question 4**

```{sql, connection=con}
SELECT 
  rating,
  AVG(replacement_cost) AS avg_replacement_cost,
  AVG(rental_rate)      AS avg_rental_rate
FROM film
GROUP BY rating;
```

# **Question 5**

```{sql, connection=con}
SELECT 
  category_id,
  COUNT(*) AS n_films
FROM film_category
GROUP BY category_id;

```

# **Question 6**

```{sql, connection=con}
SELECT 
  c.name,
  COUNT(*) AS n_films
FROM film_category AS fc
JOIN category AS c
  ON fc.category_id = c.category_id
GROUP BY c.name
ORDER BY n_films DESC
LIMIT 1;
```
